---
title: "*K. pneumoniae growth curves*"
number-sections: true
format:
  html:
    toc: true
    toc-depth: 4
editor: source
editor_options: 
  chunk_output_type: console
#bibliography: references.bib
#csl: the-american-naturalist.csl
---

```{r include = FALSE, eval = FALSE}
# The code that adds the .nojekyll file to the GitHub repository:
file <- ".nojekyll"
file.create(file)
gert::git_add(file)
gert::git_commit("Adding the .nojekyll file")
gert::git_push()
rm(file)
```

```{r include = FALSE, eval = FALSE}
run_all <- function() {
  knitr::purl("index.qmd", "tmp.R", documentation = FALSE)
  source("tmp.R")
  file.remove("tmp.R")  
}

rerun_all <- function() {
  rm(list  = grep("run_all", ls(envir = .GlobalEnv), value = TRUE, invert = TRUE),
     envir = .GlobalEnv)
  run_all()
}
```

```{r include = FALSE}
par2 <- function(...) par(..., mgp = c(1.5, .5, 0), bty = "n")

knitr::knit_hooks$set(
  margin1 = function(before, options, envir) {
    if (before) par2(plt = c(.105, .97, .15, .95)) else NULL
  })

eps <- .8
knitr::opts_chunk$set(margin1    = TRUE,
                      fig.retina = 2,
                      fig.align  = "center",
                      fig.height = eps * 5, # default is 5
                      fig.width  = eps * 7) # default is 7 and maximum possible is 8.3
```


## Constants

The files root:

```{r}
root <- paste0("/Users/", Sys.getenv("USER"), "/")
```

The mirrored folder:

```{r}
mirror_folder <- paste0(root, "GitHub/choisy/Kpneumoniae_plasmid/")
```

The root of the data folders:

```{r}
path2data <- paste0(root, "Library/CloudStorage/",
                    "OneDrive-OxfordUniversityClinicalResearchUnit/",
                    "GitHub/choisy/Kp_growth_curves/")
```

The name of the data file:

```{r}
data_file <- "CultureData_11523_Van.xlsx"
```

Defining some colors:

```{r}
col1 <- "steelblue"
col2 <- "goldenrod"
col3 <- "darkgoldenrod"
```


```{r include = FALSE}
path2cache <- paste0(path2data, "cache/")
if (! dir.exists(path2cache)) dir.create(path2cache)
make_path <- function(x) paste0(path2cache, x)
file_exists <- function(x) file.exists(make_path(x))
readRDScache <- function(x) readRDS(make_path(x))
saveRDScache <- function(object, file) saveRDS(object, make_path(file))
```


## Packages

Required packages:

```{r}
required <- c("readxl", "tidyr", "dplyr", "purrr", "stringr", "magrittr", "bbmle",
              "tibble")
```

Installing those that are not installed yet:

```{r}
to_install <- required[! required %in% installed.packages()[,"Package"]]
if (length(to_install)) install.packages(to_install)
```

Loading the packages for interactive use:

```{r warning = FALSE, message = FALSE}
invisible(sapply(required, library, character.only = TRUE))
```


## Functions

A function that excludes values from a vector `x` by their names `e`:

```{r}
exclude <- function(x, e) x[grep(e, names(x), value = TRUE, invert = TRUE)]
```

A tuning of the `plot()` function:

```{r}
plot_od <- function(..., ylab = "optical density OD600 nm") {
  plot(..., xlab = "time (hours)", ylab = ylab)
}
```

A tuning of the `bbmle::mle2()` function:

```{r}
mle3 <- function(...) mle2(..., method = "Nelder-Mead", control = list(maxit = 1e4))
```

A tuning of the `bbmle::coef()` function:

```{r}
coef2 <- function(x) exp(bbmle::coef(x))
```


## Data

Mirroring the data set if not already done:

```{r}
data_file_full <- paste0(path2data, data_file)
if (!file.exists(data_file_full))
  file.copy(paste0(mirror_folder, data_file), data_file_full)
```

Loading the data:

```{r}
(od_data <- data_file_full |>
  read_excel() |> 
  group_by(Sample, Medium) |> 
  mutate(replicate = row_number()) |>
  ungroup() |> 
  pivot_longer(- c(Sample, Medium, replicate), names_to = "time", values_to = "OD") |> 
  mutate(across(time, ~ .x |>
                  str_remove(" *h$") |> 
                  as.numeric() |> 
                  multiply_by(60) |> 
                  round() |> 
                  as.integer())))
```

Looking at the levels:

```{r}
(unique_levels <- od_data |> 
  select(-OD) |> 
  map(~ sort(unique(.x))))
```

Checking the regularity of time measurements:

```{r}
unique(diff(unique_levels$time))
```

The number of combinations:

```{r}
unique_levels |> 
  map_int(length) |> 
  reduce(`*`)
```

Number of replicates per combination:

```{r}
(nb_rep <- od_data |> 
  group_by(Sample, Medium, time) |> 
  tally())
```

Checking that we have all the combinations:

```{r}
nrow(nb_rep)
```

Looking at the number of replicates per combination:

```{r}
unique(nb_rep$n)
```

Let's now look at the data. Let's first set some colors:

```{r}
col_set <- c(col1, col2)
col_set_names <- unique_levels$Medium
col_set_alpha <- col_set |>
  adjustcolor(.3) |> 
  setNames(col_set_names)
```

A function that plots the data:

```{r}
plot_data <- function(x, y = OD, ...) {
  x |>
    mutate(across(time, ~ divide_by(.x, 60))) |> 
    slice_sample(prop = 1) |> 
    rename(y = {{ y }}) |> 
    with(plot_od(time, y, col = col_set_alpha[Medium], ...))
  legend("topleft", legend = col_set_names, pch = 1, bty = "n", col = col_set)
}
```

Plotting the OD values:

```{r plot_OD_values}
plot_data(od_data)
abline(h = c(.1, .5, .6, .8))
```

Plotting the $\Delta$OD values:

```{r plot_Delta_OD}
od_data |> 
  filter(time < 15 * 60, OD < .8) |> 
  group_by(Sample, Medium, replicate) |> 
  mutate(Delta_OD = c(NA, diff(OD))) |> 
  ungroup() |> 
  plot_data(Delta_OD, ylab = bquote( ~ Delta ~ "OD600 nm"))
```


## Probability distributions

A function that builds probability distribution functions that package everything we
need:

```{r}
distribution <- function(name, d, q) {
  attr(d, "q") <- q
  attr(d, "name") <- name
  d
}
```

Definining the normal probability distribution function:

```{r}
normal <- distribution("norm", function(x, mu, k) dnorm(x, mu, k, log = TRUE), 
                               function(x, mu, k, ...) qnorm(x, mu, k, ...))
```

Defining the log-normal probability distribution function:

```{r}
log_normal <- distribution("lnorm",
  function(x, mu, k) dlnorm(x, log(mu) - k^2 / 2, k, log = TRUE), 
  function(x, mu, k, ...) qlnorm(x, log(mu) - k^2 / 2, k, ...))
```

Defining the gamma probability distribution function:

```{r}
gammad <- distribution("gamma", function(x, mu, k) dgamma(x, k, k / mu, log = TRUE), 
                                function(x, mu, k, ...) qgamma(x, k, k / mu, ...))
```


## Growth models

Defining the exponential growth model:

```{r}
exponential_growth <- function(t, N0, r) {
  N0 * exp(r * t)
}

attr(exponential_growth, "name") <- "exponential"

exponential <- list(
  model = exponential_growth,
  
  mLL_factory = function(x, f) {
    function(N0, r, k) {
      -sum(f(x$OD, exponential_growth(x$time, exp(N0), exp(r)), exp(k)))
    }
  },
  
  ini_val = function(df, k = 1) {
    p <- setNames(c(coef(lm(log(OD) ~ time, df)), k), c("N0", "r", "k"))
    p[-1] <- log(p[-1])
    as.list(p)
  }
)
```

Defining the logistic growth model:

```{r}
logistic_growth <- function(t, K, r, t0) {
  K / (1 + exp(-(t - t0) * r))
}

attr(logistic_growth, "name") <- "logistic"

logistic <- list(
  model = logistic_growth,
  
  mLL_factory = function(x, f) {
    function(K, r, t0, k) {
      - sum(f(x$OD, logistic_growth(x$time, exp(K), exp(r), exp(t0)), exp(k)))
    }
  },
  
  ini_val = function(df, k = 1) {
    time <- df$time
    OD <- df$OD
    K <- max(OD)
    halfK <- K / 2 - min(OD) / 2                    # (second term could be optional)
    evaluate <- function(x) approx(time, OD, x)$y
    distance <- function(x) abs(evaluate(x) - halfK)
    t0 <- optimize(distance, range(time))$minimum
    r <- df |> 
      filter(time < t0) |> 
      mutate(across(OD, log)) |> 
      lm(OD ~ time, data = _) |> 
      coef() |> 
      extract(2) |> 
      unname()
    as.list(log(c(K = K, r = r, t0 = t0, k = k)))
  }
)
```

## Utility functions

A function that fits an exponential growth model by maximum-likelihood to data `x`
using a probability distribution `f`:

```{r}
calibrate <- function(x, m, f, start = NULL, k = 1, ...) {
  if (is.null(start)) start <- m$ini_val(x, k)
  out <- mle3(m$mLL_factory(x, f), start, ...)
  out@data <- list(data = x, model = m$model, distribution = f)
  out
}
```

A function that retrieves the probability distribution used from a fitted model:

```{r}
get_distr <- function(x) attr(x@data$distribution, "name")
```

A function that retrieves the growth model used from a fitted model:

```{r}
get_model <- function(x) attr(x@data$model, "name")
```

A function that retrieves the data from a fitted model:

```{r}
get_data <- function(x) x@data$data
```


```{r}
t_vals <- seq(0, 15 * 60, le = 512)

predictions <- function(x, t = t_vals, ci = .95) {
  params <- coef2(x)
  f <- attr(x@data$distribution, "q")
  q <- function(mu, lower = FALSE) f((1 - ci) / 2, mu, params["k"], lower.tail = lower)
  tibble(time = t,
         fit  = do.call(x@data$model, c(t = list(t), exclude(params, "k"))),
         lwr  = q(fit, TRUE),
         upr  = q(fit))
}
```

A function that plots the model's prediction:

```{r}
plot_predictions <- function(x, col = col1, alpha = .3, add = FALSE, ...) {
  with(x, {
    t <- time / 60
    if (! add) plot_od(t, fit, type = "n", ...)
    polygon(c(t, rev(t)), c(lwr, rev(upr)), border = NA, col = adjustcolor(col, alpha))
    lines(t, fit, col = col, lwd = 2)
  })
}
```

A function that adds data points:

```{r}
add_data <- function(x, col = col3) {
  with(get_data(x), points(time / 60, OD, col = col, pch = 19))
}
```

Let's combine them:

```{r}
plot_fit <- function(pred, fit, ylim = NULL, col_data = col3, col_model = col1,
                     alpha = .3, ...) {
  if (is.null(ylim)) {
    OD <- get_data(fit)$OD
    ylim <- c(min(c(pred$lwr, OD)), max(c(pred$upr, OD)))
  }
  plot_predictions(pred, col_model, alpha, ylim = ylim, ...)
  add_data(fit, col = col_data)
}
```

An example using a log-normal distribution:

```{r}
#logistic_fit <- od_data |> 
#  filter(Sample == "W", Medium == "M9", replicate == 1, time < 15 * 60) |> 
#  logistic_mle(log_normal)
#logistic_pred <- logistic_predict(logistic_fit)
#plot_fit(logistic_pred, logistic_fit)
```


## Calibrations

A function that fits models listed in `distr` to the (potentially grouped) data frame
`df`:

```{r}
fit_models <- function(df, distributions, model) {
  map_dfr(distributions,
          \(d) df |>
            group_modify(~ tibble(fitted = list(calibrate(.x, model, d)))) |>
            ungroup())
}
```

Let's try it:

```{r}
all_fits <- bind_rows(
              (od_data |>
                filter(between(OD, .1, .8), time < 15 * 60) |> 
                group_by(Sample, Medium, replicate) |> 
                fit_models(c(normal, log_normal, gammad), logistic)),
              (od_data |>
                filter(between(OD, .1, .8), time < 15 * 60, Medium == "M9+Fe") |>
                group_by(Sample, Medium, replicate) |> 
                fit_models(c(normal, log_normal, gammad), exponential))) |> 
  mutate(model = map_chr(fitted, get_model),
         distr = map_chr(fitted, get_distr)) |> 
  relocate(fitted, .after = last_col())
```

```{r}
all_fits
```


