---
title: "*K. pneumoniae growth curves*"
number-sections: true
format:
  html:
    toc: true
    toc-depth: 4
editor: source
editor_options: 
  chunk_output_type: console
#bibliography: references.bib
#csl: the-american-naturalist.csl
---

```{r include = FALSE, eval = FALSE}
run_all <- function() {
  knitr::purl("index.qmd", "tmp.R", documentation = FALSE)
  source("tmp.R")
  file.remove("tmp.R")  
}

add_nojekyll <- function() {
  file <- ".nojekyll"
  file.create(file)
  gert::git_add(file)
  gert::git_commit("Adding the .nojekyll file")
  gert::git_push()
  rm(file)
}
```

```{r include = FALSE}
par2 <- function(...) par(..., mgp = c(1.5, .5, 0), bty = "n")

knitr::knit_hooks$set(
  margin1 = function(before, options, envir) {
    if (before) par2(plt = c(.105, .97, .15, .95)) else NULL
  })

eps <- .8
knitr::opts_chunk$set(margin1    = TRUE,
                      fig.retina = 2,
                      fig.align  = "center",
                      fig.height = eps * 5, # default is 5
                      fig.width  = eps * 7) # default is 7 and maximum possible is 8.3
```


## Constants

The files root:

```{r}
root <- paste0("/Users/", Sys.getenv("USER"), "/")
```

The mirrored folder:

```{r}
mirror_folder <- paste0(root, "GitHub/choisy/Kpneumoniae_plasmid/")
```

The root of the data folders:

```{r}
path2data <- paste0(root, "Library/CloudStorage/",
                    "OneDrive-OxfordUniversityClinicalResearchUnit/",
                    "GitHub/choisy/Kp_growth_curves/")
```

The name of the data file:

```{r}
data_file <- "CultureData_11523_Van.xlsx"
```

```{r include = FALSE}
path2cache <- paste0(path2data, "cache/")
if (! dir.exists(path2cache)) dir.create(path2cache)
make_path <- function(x) paste0(path2cache, x)
file_exists <- function(x) file.exists(make_path(x))
readRDScache <- function(x) readRDS(make_path(x))
saveRDScache <- function(object, file) saveRDS(object, make_path(file))
```

## Packages

```{r message = FALSE}
library(readxl)
library(tidyr)
library(dplyr)
library(purrr)
library(stringr)
library(magrittr)
library(bbmle)
library(tibble)
```


## Functions

## Data

Mirroring the data set if not already done:

```{r}
data_file_full <- paste0(path2data, data_file)
if (!file.exists(data_file_full))
  file.copy(paste0(mirror_folder, data_file), data_file_full)
```

Loading the data:

```{r}
(od_data <- data_file_full |>
  read_excel() |> 
  group_by(Sample, Medium) |> 
  mutate(replicate = row_number()) |>
  ungroup() |> 
  pivot_longer(- c(Sample, Medium, replicate), names_to = "time", values_to = "OD") |> 
  mutate(across(time, ~ .x |>
                  str_remove(" *h$") |> 
                  as.numeric() |> 
                  multiply_by(60) |> 
                  round() |> 
                  as.integer())))
```

Looking at the levels:

```{r}
(unique_levels <- od_data |> 
  select(-OD) |> 
  map(~ sort(unique(.x))))
```

Checking the regularity of time measurements:

```{r}
unique(diff(unique_levels$time))
```

The number of combinations:

```{r}
unique_levels |> 
  map_int(length) |> 
  reduce(`*`)
```

Number of replicates per combination:

```{r}
(nb_rep <- od_data |> 
  group_by(Sample, Medium, time) |> 
  tally())
```

Checking that we have all the combinations:

```{r}
nrow(nb_rep)
```

Looking at the number of replicates per combination:

```{r}
unique(nb_rep$n)
```


## Plots

Setting the colors:

```{r}
col_set <- c("steelblue", "goldenrod")
col_set_names <- unique_levels$Medium
col_set_alpha <- col_set |>
  adjustcolor(.3) |> 
  setNames(col_set_names)
```

A function that plots the data:

```{r}
plot_data <- function(x, y = OD, ylab = "optical density OD600 nm") {
  x |>
    mutate(across(time, ~ divide_by(.x, 60))) |> 
    slice_sample(prop = 1) |> 
    rename(y = {{ y }}) |> 
    with(plot(time, y, col = col_set_alpha[Medium],
              xlab = "time (hours)", ylab = ylab))
  legend("topleft", legend = col_set_names, pch = 1, bty = "n", col = col_set)
}
```

Plotting the OD values:

```{r plot_OD_values}
plot_data(od_data)
abline(h = c(.1, .8))
```

Plotting the $\Delta$OD values:

```{r plot_Delta_OD}
od_data |> 
  filter(time < 15 * 60, OD < .8) |> 
  group_by(Sample, Medium, replicate) |> 
  mutate(Delta_OD = c(NA, diff(OD))) |> 
  ungroup() |> 
  plot_data(Delta_OD, ylab = "Delta OD600 nm")
```

## Models

The logistic growth model:

```{r}
logistic <- function(t, K, r, t0) {
  K / (1 + exp(-(t - t0) * r))
}
```

A function that adds metadata to a function:

```{r}
add_metadata <- function(f, name, q) {
  attr(f, "name") <- name
  attr(f, "q") <- q
  f
}
```

A reparameterization of `qgamma()`:

```{r}
qgamma2 <- function(x, mu, k, ...) {
    qgamma(x, k, k / mu, ...)
}
```

A reparameterization of `dgamma()`:

```{r}
dgamma2 <- add_metadata(function(x, mu, k) dgamma(x, k, k / mu, log = TRUE),
                        "gamma", qgamma2)
```

A reparameterization of `qlnorm()`:

```{r}
qlnorm2 <- function(x, mu, k, ...) {
    qlnorm(x, log(mu), k, ...)
}
```

A reparameterization of `dlnorm()`:

```{r}
dlnorm2 <- add_metadata(function(x, mu, k) dlnorm(x, log(mu), k, log = TRUE),
                        "lnorm", qlnorm2)
```

The minus log-likelihood function factory:

```{r}
make_mLL <- function(x, f) {
  function(K, r, t0, k) {
    - sum(f(x$OD, logistic(x$time, exp(K), exp(r), exp(t0)), exp(k)))  
  }
}
```

A function that looks for good initial values for the parameters of the logistic growth
model:

```{r}
ini_val <- function(df, k = 1) {
  time <- df$time
  K <- max(df$OD)
  halfK <- K / 2
  evaluate <- function(x) approx(time, df$OD, x)$y
  distance <- function(x) abs(evaluate(x) - halfK)
  t0 <- optimize(distance, range(time))$minimum
  r <- df |> 
    filter(time < t0) |> 
    mutate(across(OD, log)) |> 
    lm(OD ~ time, data = _) |> 
    coef() |> 
    extract(2) |> 
    unname()
  as.list(log(c(K = K, r = r, t0 = t0, k = k)))
}
```

A tuning of the `mle2()` function that includes the automatic initial values of the 
parameters search and the specification of the density of probability we want to use
for the minus log-likelihood function:

```{r}
mle3 <- function(x, f, k = 1, ...) {
  out <- mle2(make_mLL(x, f), ini_val(x, k), "Nelder-Mead",
              control = list(maxit = 1e4), ...)
  out@data <- list(data = x, distribution = f)
  out
}
```

An example using a log-normal distribution:

```{r}
od_data |> 
  filter(Sample == "W", Medium == "M9", replicate == 1, time < 15 * 60) |> 
  mle3(dlnorm2)
```


## Batches

A function that fits a model on all the replicates of a data frame:

```{r}
fit_by_replicate <- function(x, f) {
  x |> 
    group_by(Sample, replicate) |> 
    group_split() |> 
    map(mle3, f)
}
```

Fitting a log-normal likelihood to the replicates from the M9 medium:

```{r}
M9_medium_lnorm <- od_data |> 
  filter(Medium == "M9", time < 15 * 60) |>
  fit_by_replicate(dlnorm2)
```

A function that summarizes the fits of an output from `fit_by_replicate()`:

```{r}
fits_summary <- function(x) {
  bind_cols(map_dfr(x, ~ tibble(dist = attr(.x@data$distribution, "name"))),
            map_dfr(x, ~ unique(select(.x@data$data, -time, -OD))),
            map_dfr(x, ~ as_tibble_row(exp(coef(.x)))),
            map_dfr(x, ~ tibble(LL = as.numeric(logLik(.x)))),
            map_dfr(x, ~ tibble(conv = .x@details$convergence)))
}
```

Looking at the summary of the fits on the replicates from M9 medium using a log-normal
likelihood:

```{r}
M9_medium_lnorm |>
  fits_summary() |> 
  print(n = Inf)
```

A function that performs model predictions:

```{r}
predictions <- function(x, t, ci = .95) {
  params <- as.list(exp(coef(x)))
  f <- attr(x@data$distribution, "q")
  q <- function(meanlog, lower = FALSE) f(1 - ci / 2, meanlog, params$k, lower)
  tibble(time = t,
         fit = with(params, logistic(t, K, r, t0)),
         lwr = q(fit),
         upr = q(fit, TRUE))
}
```

Let's try it:

```{r}
a <- predictions(M9_medium_lnorm[[1]], seq(0, 15 * 60, le = 512))
```

A function that plots the model's prediction:

```{r}
plot_predictions <- function(x, col = "steelblue", alpha = .3) {
  with(x, {
    t <- time / 60
    plot(t, fit, type = "n", xlab = "time (hours)", ylab = "OD 600nm")
    polygon(c(t, rev(t)), c(lwr, rev(upr)), border = NA, col = adjustcolor(col, alpha))
    lines(t, fit, col = col)
  })
}
```

Let's try it:

```{r}
plot_predictions(a)
```


