---
title: "*K. pneumoniae growth curves*"
number-sections: true
format:
  html:
    toc: true
    toc-depth: 4
editor: source
editor_options: 
  chunk_output_type: console
#bibliography: references.bib
#csl: the-american-naturalist.csl
---

```{r include = FALSE, eval = FALSE}
run_all <- function() {
  knitr::purl("index.qmd", "tmp.R", documentation = FALSE)
  source("tmp.R")
  file.remove("tmp.R")  
}

add_nojekyll <- function() {
  file <- ".nojekyll"
  file.create(file)
  gert::git_add(file)
  gert::git_commit("Adding the .nojekyll file")
  gert::git_push()
  rm(file)
}
```

```{r include = FALSE}
par2 <- function(...) par(..., mgp = c(1.5, .5, 0), bty = "n")

knitr::knit_hooks$set(
  margin1 = function(before, options, envir) {
    if (before) par2(plt = c(.105, .97, .15, .95)) else NULL
  })

eps <- .8
knitr::opts_chunk$set(margin1    = TRUE,
                      fig.retina = 2,
                      fig.align  = "center",
                      fig.height = eps * 5, # default is 5
                      fig.width  = eps * 7) # default is 7 and maximum possible is 8.3
```


## Constants

```{r}
root <- paste0("/Users/", Sys.getenv("USER"), "/")
```

The mirrored folder:

```{r}
mirror_folder <- paste0(root, "GitHub/choisy/Kpneumoniae_plasmid/")
```

The root of the data folders:

```{r}
path2data <- paste0(root, "Library/CloudStorage/",
                    "OneDrive-OxfordUniversityClinicalResearchUnit/",
                    "GitHub/choisy/Kp_growth_curves/")
```

The name of the data file:

```{r}
data_file <- "CultureData_11523_Van.xlsx"
```

```{r include = FALSE}
path2cache <- paste0(path2data, "cache/")
if (! dir.exists(path2cache)) dir.create(path2cache)
make_path <- function(x) paste0(path2cache, x)
file_exists <- function(x) file.exists(make_path(x))
readRDScache <- function(x) readRDS(make_path(x))
saveRDScache <- function(object, file) saveRDS(object, make_path(file))
```

## Packages

```{r}
# stringr
```


```{r message = FALSE}
library(readxl)
library(tidyr)
library(dplyr)
library(purrr)
library(magrittr)
```


## Functions

## Data

Mirroring the data set if not already done:

```{r}
data_file_full <- paste0(path2data, data_file)
if (!file.exists(data_file_full))
  file.copy(paste0(mirror_folder, data_file), data_file_full)
```

Loading the data:

```{r}
(od_data <- data_file_full |>
  read_excel() |> 
  group_by(Sample, Medium) |> 
  mutate(replicate = row_number()) |>
  ungroup() |> 
  pivot_longer(- c(Sample, Medium, replicate), names_to = "time", values_to = "OD") |> 
  mutate(across(time, ~ .x |>
                  stringr::str_remove(" *h$") |> 
                  as.numeric() |> 
                  multiply_by(60) |> 
                  round() |> 
                  as.integer())))
```

Looking at the levels:

```{r}
(unique_levels <- od_data |> 
  select(-OD) |> 
  map(~ sort(unique(.x))))
```

Checking the regularity of time measurements:

```{r}
unique(diff(unique_levels$time))
```

The number of combinations:

```{r}
unique_levels |> 
  map_int(length) |> 
  reduce(`*`)
```

Number of replicates per combination:

```{r}
(nb_rep <- od_data |> 
  group_by(Sample, Medium, time) |> 
  tally())
```

Checking that we have all the combinations:

```{r}
nrow(nb_rep)
```

Looking at the number of replicates per combination:

```{r}
unique(nb_rep$n)
```


## Plots

```{r}
col_set <- c("steelblue", "goldenrod")
col_set_names <- unique_levels$Medium
col_set_alpha <- col_set |>
  adjustcolor(.3) |> 
  setNames(col_set_names)

od_data |>
  mutate(across(time, ~ divide_by(.x, 60))) |> 
  slice_sample(prop = 1) |> 
  with(plot(time, OD, col = col_set_alpha[Medium],
            xlab = "time (hours)", ylab = "optical density OD600 nm"))

legend("topleft", legend = col_set_names, pch = 1, bty = "n", col = col_set)
```

```{r}
od_data |> 
  filter(time < 15 * 60)
```

